Follow-up Samples

### 2017-01-15 submission/manuscript_outline_20161024.Rmd
updated summary added to discussion

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was no significant difference between initial and follow up samples for any single OTU [Figure S1].

@@

***Full and Reduced Model Results***
Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.  Interestingly, identification of the most important variables and reducing the variables considered to only these increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).

@@

**Treatment and Time Differences**
There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

@@

## Discussion
In our training set we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1].  With respect to our test set there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma [Figure 1a].  In contrast, FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].  An NMDS showed that there was very little observable change between initial and follow up for the adenoma group but there was one for the carcinoma group [Figure 1c & 1d].  This cursory information is suggestive that treatment of carcinoma, had the largest response.

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r #format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r #format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r #followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r #sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].  Many of the OTUs identified classified to normal flora bacterium [Table S3].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

>>>>

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery were visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was no significant difference between initial and follow up samples for any single OTU [Figure S1].

@@

***Full and Reduced Model Results***
Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.  Interestingly, identification of the most important variables and reducing the models to only these factors increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).

@@

***Treatment and Time Differences***
There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).  Time elapsed between the collection of the follow up sample from the initial sample, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

@@

## Discussion
There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].  Yet based on thetayc there was a significant difference between inital and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.  The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].  There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].  Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].

Interestingly, when previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or chaning due to surgery.  We found that the OTUs that made up most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from previously associated cancer bacteria (Porphyromonas asaccharolytica).  Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and all belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described on specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
